Reanalysis Step 6: Alternative Propensity Score Specifications

### 2025-08-29 reanalysis-3-imputation_diagnostics.qmd
wip

    # Lapse status
    pct_lapsed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),

@@

@@

@@

@@

@@

  cat("\n### Recommendations:\n\n")
  cat("Consider alternative imputation approaches that:\n")
  cat("1. Incorporate temporal constraints (e.g., monotone imputation)\n")
  cat("2. Respect clinical boundaries on plausible changes\n")
  cat("3. Use the intermediate values more directly when available\n")
  cat("4. Consider parametric models that can incorporate domain knowledge about vision progression\n")
} else {
  cat("\n### Conclusion:\n")
  cat("The imputation procedure appears to be producing clinically plausible values that are similar to observed outcomes.\n")
}

>>>>

    # Lapse status - distinguish between observed and missing
    n_lapse_observed = sum(!is.na(person_ever_lapse)),
    n_lapse_missing = sum(is.na(person_ever_lapse)),
    pct_lapse_missing = round(100 * n_lapse_missing / n(), 1),
    # Among those with observed lapse status
    pct_lapsed_observed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),

@@

## Lapse Status Missingness Analysis

```{r lapse-missingness}
# Detailed analysis of lapse status by group
lapse_analysis <- df_original |>
  group_by(imputation_group) |>
  summarise(
    n_total = n(),

    # Lapse status availability
    n_lapse_known = sum(!is.na(person_ever_lapse)),
    n_lapse_missing = sum(is.na(person_ever_lapse)),
    pct_lapse_missing = round(100 * n_lapse_missing / n_total, 1),

    # Among those with known lapse status
    n_lapsed_if_known = sum(person_ever_lapse == TRUE, na.rm = TRUE),
    n_not_lapsed_if_known = sum(person_ever_lapse == FALSE, na.rm = TRUE),
    pct_lapsed_if_known = round(100 * n_lapsed_if_known / n_lapse_known, 1),

    # Overall lapse rate (including imputed - ever_lapse_binary should be complete)
    pct_lapsed_overall = round(100 * mean(ever_lapse_binary == 1), 1),

kable(lapse_analysis,
      col.names = c("Group", "N Total", "N Lapse Known", "N Lapse Missing", 
                    "% Lapse Missing", "N Lapsed (if known)", "N Not Lapsed (if known)",
                    "% Lapsed (if known)", "% Lapsed (overall)"),
      caption = "Lapse status availability and rates by imputation group")

# Cross-tabulation of outcome availability vs lapse status
outcome_by_lapse <- df_original |>
  mutate(
    has_outcome = !is.na(outcome_VA_logMAR),
    lapse_known = !is.na(person_ever_lapse)
  ) |>
  group_by(lapse_known, has_outcome) |>
  summarise(
    n = n(),
    pct_of_total = round(100 * n / nrow(df_original), 1),
    # Among those with known lapse status
    pct_lapsed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(outcome_by_lapse,
      caption = "Relationship between outcome availability and lapse status knowledge")

# Key insight: Check if missing outcomes correlate with lapse
missing_outcome_lapse <- df_original |>
  filter(!is.na(person_ever_lapse)) |>  # Only those with known lapse status
  group_by(ever_lapse_binary) |>
  summarise(
    n = n(),
    n_missing_outcome = sum(is.na(outcome_VA_logMAR)),
    pct_missing_outcome = round(100 * n_missing_outcome / n, 1),
    .groups = "drop"
  ) |>
  mutate(
    lapse_status = ifelse(ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
  ) |>
  select(lapse_status, n, n_missing_outcome, pct_missing_outcome)

kable(missing_outcome_lapse,
      col.names = c("Lapse Status", "N", "N Missing Outcome", "% Missing Outcome"),
      caption = "Outcome missingness by known lapse status")
```

@@

## Clinical Trajectory by Lapse Status

```{r trajectory-by-lapse}
# Analyze vision changes by lapse status (across all imputation groups)
trajectory_by_lapse <- changes_df |>
  filter(!is.na(change_baseline_to_outcome)) |>
  group_by(ever_lapse_binary, imputation_group) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    pct_improved = round(100 * mean(outcome_direction == "Improved", na.rm = TRUE), 1),
    pct_stable = round(100 * mean(outcome_direction == "Stable", na.rm = TRUE), 1),
    pct_worsened = round(100 * mean(outcome_direction == "Worsened", na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  mutate(
    lapse_status = ifelse(ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
  ) |>
  select(-ever_lapse_binary) |>
  arrange(imputation_group, lapse_status)

kable(trajectory_by_lapse |>
        mutate(across(where(is.numeric) & !n, ~round(., 3))),
      caption = "Vision trajectories by lapse status and imputation group")

# Overall comparison (pooled across groups)
overall_by_lapse <- changes_df |>
  filter(!is.na(change_baseline_to_outcome)) |>
  group_by(ever_lapse_binary) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    pct_improved = round(100 * mean(outcome_direction == "Improved", na.rm = TRUE), 1),
    pct_worsened = round(100 * mean(outcome_direction == "Worsened", na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  mutate(
    lapse_status = ifelse(ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
  ) |>
  select(lapse_status, everything(), -ever_lapse_binary)

kable(overall_by_lapse |>
        mutate(across(where(is.numeric) & !n, ~round(., 3))),
      caption = "Overall vision trajectories by lapse status (pooled across all groups)")

# Statistical test
lapse_test <- t.test(change_baseline_to_outcome ~ ever_lapse_binary, data = changes_df)

cat("T-test comparing vision changes between lapsed and not lapsed patients:\n")
cat(sprintf("Mean difference: %.3f logMAR units\n", diff(lapse_test$estimate)))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", lapse_test$conf.int[1], lapse_test$conf.int[2]))
cat(sprintf("p-value: %.4f\n", lapse_test$p.value))

# Visualization
p_lapse <- ggplot(changes_df |> filter(!is.na(change_baseline_to_outcome)),
                  aes(x = factor(ever_lapse_binary, labels = c("Not Lapsed", "Lapsed")),
                      y = change_baseline_to_outcome,
                      fill = factor(ever_lapse_binary))) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~imputation_group) +
  scale_fill_manual(values = c("0" = "#2E86AB", "1" = "#A23B72")) +
  labs(
    title = "Vision Changes by Lapse Status Across Imputation Groups",
    subtitle = "Positive values indicate vision worsening",
    x = "Lapse Status",
    y = "Change in logMAR"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p_lapse)

# Check if lapsed patients have paradoxically better outcomes
paradox_check <- trajectory_by_lapse |>
  group_by(imputation_group) |>
  summarise(
    lapsed_mean = mean_change[lapse_status == "Lapsed"],
    not_lapsed_mean = mean_change[lapse_status == "Not Lapsed"],
    difference = lapsed_mean - not_lapsed_mean,
    paradoxical = difference < 0,  # Negative means lapsed have better outcomes
    interpretation = ifelse(
      difference < -0.05, "Lapsed have notably better vision",
      ifelse(difference > 0.05, "Lapsed have worse vision (expected)",
             "Similar outcomes")
    ),
    .groups = "drop"
  )

kable(paradox_check,
      caption = "Clinical plausibility check: Do lapsed patients have worse vision outcomes?")
```

@@

### Key Insights on Lapse Status and Outcomes

```{r lapse-insights}
# Summarize key findings about lapse status
cat("#### Lapse Status Missingness:\n\n")

# Check if lapse status is mostly observed
lapse_missing_overall <- sum(is.na(df_original$person_ever_lapse)) / nrow(df_original) * 100
if (lapse_missing_overall < 5) {
  cat(sprintf("- Lapse status is well-documented (only %.1f%% missing)\n", lapse_missing_overall))
} else {
  cat(sprintf("- ⚠ Lapse status has %.1f%% missing values\n", lapse_missing_overall))
}

# Report on high lapse rates in imputed groups
cat("\n#### Lapse Rates by Group:\n\n")
for (i in 1:nrow(lapse_analysis)) {
  group <- lapse_analysis$imputation_group[i]
  lapse_rate <- lapse_analysis$pct_lapsed_overall[i]
  cat(sprintf("- %s: %.1f%% lapsed\n", group, lapse_rate))
}

cat("\n#### Clinical Trajectory Plausibility:\n\n")
# Check the paradox
for (i in 1:nrow(paradox_check)) {
  group <- paradox_check$imputation_group[i]
  interp <- paradox_check$interpretation[i]
  diff <- paradox_check$difference[i]
  cat(sprintf("- %s: %s (difference = %.3f logMAR)\n", group, interp, diff))
}

cat("\n#### Interpretation:\n\n")
# High lapse rate interpretation
if (any(lapse_analysis$pct_lapsed_overall[lapse_analysis$imputation_group != "Observed"] > 95)) {
  cat("The extremely high lapse rates (>95%) in some imputed groups may actually be accurate:\n")
  cat("- These patients lack 2-year outcomes precisely BECAUSE they lapsed from care\n")
  cat("- The missingness mechanism is directly related to the lapse status\n")
  cat("- This is a form of 'Missing Not At Random' (MNAR) where missingness itself is informative\n\n")
}

# Clinical trajectory interpretation
if (any(paradox_check$paradoxical)) {
  cat("⚠ CONCERN: Lapsed patients show better vision outcomes in some groups:\n")
  cat("- This violates clinical expectations (lapsed patients typically have worse outcomes)\n")
  cat("- Possible explanations:\n")
  cat("  1. Selection bias: Healthier patients may be more likely to skip appointments\n")
  cat("  2. Imputation artifact: The model may not properly account for the lapse-outcome relationship\n")
  cat("  3. Data issue: Lapse definition or timing may not align with clinical deterioration\n\n")
}
```

@@

# Add new concern about lapse-outcome paradox
if (any(paradox_check$paradoxical)) {
  concerns <- c(concerns,
    "Lapsed patients show better vision outcomes than non-lapsed patients in some groups, which is clinically implausible.")
}

# Add concern about extreme lapse rates
if (any(lapse_analysis$pct_lapsed_overall[lapse_analysis$imputation_group != "Observed"] > 95)) {
  concerns <- c(concerns,
    "Extremely high lapse rates (>95%) in imputed groups suggest the imputation may be conflating missingness with lapse status.")
}

@@

  cat("\n### Recommendations:\n\n")
  cat("Consider the following approaches:\n\n")
  cat("1. **Sensitivity Analysis**: Compare results using:\n")
  cat("   - Complete case analysis only\n")
  cat("   - Imputation restricted to those with intermediate values\n")
  cat("   - Different assumptions about lapsed patient outcomes\n\n")
  cat("2. **Alternative Imputation Methods**:\n")
  cat("   - Use pattern mixture models that explicitly model different trajectories for lapsed vs non-lapsed\n")
  cat("   - Implement bounded imputation that restricts vision changes to clinically plausible ranges\n")
  cat("   - Consider worst-case imputation for lapsed patients (assume deterioration)\n\n")
  cat("3. **Reframe the Analysis**:\n")
  cat("   - Treat missing outcomes as a competing risk rather than missing data\n")
  cat("   - Focus on the population that remained in care (per-protocol analysis)\n")
  cat("   - Use inverse probability weighting to account for differential dropout\n\n")
} else {
  cat("\n### Conclusion:\n")
  cat("The imputation procedure appears to be producing clinically plausible values that are similar to observed outcomes.\n")
}

### 2025-08-29 reanalysis-5-outcome_analysis-interaction.qmd
wip

```{r setup}
#| include: false
source("dependencies.R")
setup_analysis(seed = 2025)

@@

```{r define-formula}
# Base covariates (same as main analysis)
base_covariates <- c(
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat", 
  "insurance_cat",
  "age_cat",
  "CCI",
  "DCSI",
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat"
)

# Three-way interaction formula
# This creates all main effects, two-way interactions, and the three-way interaction
interaction_formula <- as.formula(paste(
  "outcome_va_vi_binary ~ ever_lapse_binary * dr_severity * any_treatment +",
  paste(base_covariates, collapse = " + ")
))

>>>>

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")  # Load formulas and helper functions
setup_analysis(seed = 2025)

@@

```{r define-formula}
# Get the interaction formulas from functions.R
interaction_formulas <- get_analysis_formulas_reanalysis_interaction()

# Use the full formula which includes three-way interaction
interaction_formula <- interaction_formulas$full

### 2025-08-29 reanalysis-5-outcome_analysis.qmd
wip

```{r fit-models}
# Check if models are already cached
models_cache_file <- file.path(reanalysis_data_dir, "fitted_models_main_effects.rds")
force_refit <- FALSE  # Set to TRUE if you want to force re-fitting

>>>>

```{r fit-models}
# Check if models are already cached
models_cache_file <- file.path(reanalysis_data_dir, "fitted_models_main_effects.rds")
force_refit <- TRUE  # Set to TRUE if you want to force re-fitting

### 2025-08-29 reanalysis-7-alternative_specifications.qmd
wip

```{r fit-outcomes}
# Define outcome model formula (main effects only)
base_covariates <- c(
  "baseline_VA_logMAR", "gender_cat", "race_ethnic_cat", 
  "insurance_cat", "age_cat", "CCI", "DCSI",
  "glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat"
)

outcome_formula <- as.formula(paste(
  "outcome_va_vi_binary ~ ever_lapse_binary + dr_severity + any_treatment +",
  paste(base_covariates, collapse = " + ")
))

>>>>

```{r fit-outcomes}
# Get the interaction formulas from functions.R
interaction_formulas <- get_analysis_formulas_reanalysis_interaction()

# Use the full formula which includes three-way interaction
outcome_formula <- interaction_formulas$full
